---
title: "BBRC POMP calibration results"
output:
  html_document:
    theme: united
    highlight: tango
    code_folding: hide
  html_notebook:
     code_folding: hide
  pdf_document: default
  
---

```{r echo=FALSE, message = FALSE, include=FALSE}
library(ggfortify)
library(tidyverse)
library(glue)
source(file = 'COVID-pomp/scripts/utils.R')
```

Launched 4 calibrations:

 - **1_3**: One compartments for $E$ and three for $I$, control experiment, as swiss medical weekly paper.
 - **9_14**: Nine compartments for $E$ and fourteen for $I$, intermediate experiment.
 - **70_100**: 70 compartments for $E$ and 100 for $I$, much closer to a constant serial interval.
 - **dirac**: deterministic true dirac forcing the outing of the $E$ compartments after 2.9 days and  the outing of the $I$ compartement after 4.6 days. The dirac is precise at 1/10th of a day.


```{r echo=FALSE, message = FALSE}
filter1_3 <- readRDS(glue("COVID-pomp/results/echopc16/first1-3/COVID-pomp/results/filtered_COVID_CH_CH_d-deltah_R0_0-I_0-id2o_30_NB_1_3.rds")) %>%   
  group_by(time, parset, var) %>% 
  summarise(mean = mean(value, na.rm = T),
            median = median(value, na.rm = T),
            q025 = quantile(value, 0.025, na.rm = T),
            q975 = quantile(value, 0.975, na.rm = T),
            q25 = quantile(value, 0.25, na.rm = T),
            q75 = quantile(value, 0.75, na.rm = T)) %>%
  ungroup() %>% 
  mutate(date = yearsToDate(time),
         parset = factor(parset)) %>% 
  filter(var == 'Rt') %>%
  select(-parset, -var) %>% 
  mutate(params = '1-3')

filter9_14 <- readRDS(glue("COVID-pomp/results/echopc16/first9-14/COVID-pomp/results/filtered_COVID_CH_CH_d-deltah_R0_0-I_0-id2o_30_NB_9_14.rds")) %>% 
  group_by(time, parset, var) %>% 
  summarise(mean = mean(value, na.rm = T),
            median = median(value, na.rm = T),
            q025 = quantile(value, 0.025, na.rm = T),
            q975 = quantile(value, 0.975, na.rm = T),
            q25 = quantile(value, 0.25, na.rm = T),
            q75 = quantile(value, 0.75, na.rm = T)) %>%
  ungroup() %>% 
  mutate(date = yearsToDate(time),
         parset = factor(parset)) %>% 
  filter(var == 'Rt') %>%
  select(-parset, -var) %>% 
  mutate(params = '9-14')

filter_dirac <- readRDS(glue("COVID-pomp/results/echopc16/dirac/COVID-pomp/results/filtered_COVID_CH_CH_d-deltah_R0_0-I_0-id2o_30_NB_29_46.rds")) %>%   
  group_by(time, parset, var) %>% 
  summarise(mean = mean(value, na.rm = T),
            median = median(value, na.rm = T),
            q025 = quantile(value, 0.025, na.rm = T),
            q975 = quantile(value, 0.975, na.rm = T),
            q25 = quantile(value, 0.25, na.rm = T),
            q75 = quantile(value, 0.75, na.rm = T)) %>%
  ungroup() %>% 
  mutate(date = yearsToDate(time),
         parset = factor(parset)) %>% 
  filter(var == 'Rt') %>%
  select(-parset, -var) %>% 
  mutate(params = 'dirac')

filter70_100 <- readRDS(glue("COVID-pomp/results/echopc16/first70-100/COVID-pomp/results/filtered_COVID_CH_CH_d-deltah_R0_0-I_0-id2o_30_NB_70_100.rds")) %>%   
  group_by(time, parset, var) %>% 
  summarise(mean = mean(value, na.rm = T),
            median = median(value, na.rm = T),
            q025 = quantile(value, 0.025, na.rm = T),
            q975 = quantile(value, 0.975, na.rm = T),
            q25 = quantile(value, 0.25, na.rm = T),
            q75 = quantile(value, 0.75, na.rm = T)) %>%
  ungroup() %>% 
  mutate(date = yearsToDate(time),
         parset = factor(parset)) %>% 
  filter(var == 'Rt') %>%
  select(-parset, -var) %>% 
  mutate(params = '70-100')

filter_all <- rbind(filter1_3, filter9_14, filter70_100, filter_dirac)
```




```{r fig.cap="R0 infered using the different parametrisations."}

p <- ggplot(filter_all, 
            aes(x = time)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = .2) +
  geom_ribbon(aes(ymin = q25, ymax = q75), alpha = .2) +
  facet_wrap(~params)  +
  theme_bw() +
  scale_x_continuous(labels = yearsToDateLabel)
p
```

```{r fig.cap="R0 infered using different parametrisations."}
p <- ggplot(filter_all, 
            aes(x = time)) +
  geom_ribbon(aes(ymin = q025, ymax = q975, fill = params), alpha = .1) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = params), alpha = .3) +
  #geom_line(aes(y = median, fill = params)) +
  theme_bw() +
  scale_x_continuous(labels = yearsToDateLabel)
p
```
```{r fig.cap="R0 infered using different parametrisations maybe clearer with just q25->175"}
p <- ggplot(filter_all, 
            aes(x = time)) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = params), alpha = .2) +
  geom_line(aes(y = median, color = params)) +
  theme_bw() +
  scale_x_continuous(labels = yearsToDateLabel)
p
```

# Appendix: distributions

```{r fig.cap="Distribution of time spent infectious: mean is 4.6 days, the dirac is shown in dashed red, gamma with shape 3 in green, with shape 14 in blue and with shape 100 in orange"}
shape <- 3
p <- ggdistribution(dgamma, seq(0, 10, 0.1), shape = shape, scale= 4.6/shape, c = 'green')
shape <- 14
p <- ggdistribution(dgamma, seq(0, 10, 0.1), shape = shape, scale= 4.6/shape, p = p, c= 'blue')
shape <- 100
p <- ggdistribution(dgamma, seq(0, 10, 0.1), shape = shape, scale= 4.6/shape, p = p, c= 'orange')
p + geom_vline(xintercept=4.6-0.05, linetype="dashed", color = "red") +geom_vline(xintercept=4.6+0.05, linetype="dashed", color = "red")
```

```{r fig.cap="Distribution of time spent incubating: mean is 2.9 days, , the dirac is shown in dashed red, gamma with shape 1 in green, with shape 9 in blue and with shape 70 in orange."}

shape <- 1
p <- ggdistribution(dgamma, seq(0, 8, 0.1), shape = shape, scale= 2.9/shape, c = 'green')
shape <- 9
p <- ggdistribution(dgamma, seq(0, 8, 0.1), shape = shape, scale= 2.9/shape, p = p, c= 'blue')
shape <- 70
p <- ggdistribution(dgamma, seq(0, 8, 0.1), shape = shape, scale= 2.9/shape, p = p, c= 'orange')
p + geom_vline(xintercept=2.9-0.05, linetype="dashed", color = "red") + geom_vline(xintercept=2.9+0.05, linetype="dashed", color = "red")
```